Point-process based representation learning for Electronic Health Records

### 2023-03-03 Manuscript_v1.tex
update
update ol

\usepackage{svg}

@@

    Irregular sampling of time series in electronic health records (EHRs) is challenging for model development. In addition, the pattern of missingness for certain clinical variables are not at random as it is determined by clinicians decision and the state of patient. Point process is a mathematical framework for handling event sequence data which is also consistent with the irregular sampling. We propose \emph{TEEDAM}, a deep neural network that can learn patient representation from irregulary sample time series as well as informative missingness pattern of certain laboratory variables. We performed various experiments to show the effectiveness of event and state encoding for characterization of conditional intensity functions as well as downstream prediction task. Results show that in some cases learning from patterns may improve the performance of prediction task.
\end{abstract}

@@

[ML in healthcare]Machine learning has the potential to revolutionize healthcare by leveraging vast amounts of data available in electronic health records (EHRs) to develop more accurate clinical decision support systems. EHRs store patient health information, such as medical history, medications, lab results, and diagnostic images, which can be used as input for machine learning algorithms to identify patterns and associations that could inform more precise diagnoses, better treatment plans, and earlier interventions. Clinical decision support systems that use machine learning can provide real-time, evidence-based recommendations to healthcare providers, reducing errors and improving patient outcomes.

[irregular sampling challenge] One of data challenges for machine learning (ML) when using electronic health records (EHRs) is irregular sampling. EHR data is often collected at different times and frequencies, depending on a patient's healthcare needs and visit schedules, which can result in uneven and irregularly sampled time series data.

@@

[sources of missingness] Sources of missingness in EHRs must be carefully understood. For example, lab measurements are typically ordered as part of a diagnostic work-up, meaning that the presence of a datapoint conveys information about the patient's state
% \cite*{ghassemiReviewChallengesOpportunities2020}.

[conseqs of IRR] A typical EHR consist of tabular and time series data. Depending on the desired outcome, we can formulate the problem in static or longitudinal setting. There are various consequences asscociated with irregularly sampled data. Many machine learning models are not inherently compatible with

imputation for static setting

imputation for time series.

inherently compatible methods.

imputation methods/// Imputation techniques are widely used in electronic health records (EHRs) to handle missing data. Imputation refers to the process of filling in the missing data with plausible values based on the available information. There are several imputation techniques available, each with their own strengths and weaknesses. For example, mean imputation replaces missing values with the mean of the available values in the same column, while regression imputation uses a regression model to estimate the missing values based on the relationship between the variables. Other popular imputation methods include multiple imputation, hot deck imputation, and k-nearest neighbors imputation. The choice of imputation technique depends on the type and amount of missing data, as well as the goals of the analysis. Regardless of the technique used, it is important to carefully consider the impact of imputed values on the analysis and to report the imputation method used in the results.

There exist several ML frameworks that are inherently compatible with the missig or irregularly-sampled data such as Gaussian process and recurrent neural networks.
Deep learning models that are compatible with irregularly sampled time series are important in many real-world applications, such as stock market predictions, speech recognition, and medical diagnosis. These models need to be able to handle data that is not evenly spaced, as is often the case in time-sensitive applications. One popular deep learning architecture that can handle irregularly sampled time series is the Recurrent Neural Network (RNN). RNNs are well-suited for this task because they have the ability to process sequential data, taking into account not only the current input but also the previous inputs. Another architecture that can handle irregularly sampled time series is the Convolutional Neural Network (CNN), which can be used to extract features from the data and then pass the processed data to an RNN for further analysis. Additionally, Attention Mechanisms can be integrated into RNNs and CNNs to help the model focus on important features in the data. Overall, these deep learning models offer a powerful toolset for handling irregularly sampled time series and provide useful insights into this type of data.
These models does not explicitly take into account the fact wether the absence of some datapoints could convey some unique information.

Point processes are mathematical models used to describe the distribution of events in time or space. These events can be occurrences of earthquakes, nerve impulses, customer purchases, or anything else that can be counted or measured.
In healthcare, lab measurements can be regarded as sequence of events that are ordered by clinicians. At the core of point process is the defintion of conditional intensity function and the corresponding log likelihood which simultanously models the occurence of events using the history of past events.[GAP]
More recently, neural Point Processes have been developed for better characteriztion of CIFs by leveraing the power of deep neural networks. These models can be used for tasks such as predicting future events, estimating the rate of event occurrence, or identifying correlations between events. They offer a flexible and powerful way to analyze point process data, as they can handle complex dependencies between events and incorporate prior knowledge about the process. Traditioaly, a point process encodes timestample and type of events, however, it is common in healtcare to have additional sources of information that could be useful for CIF characterization. For example the absolute value of patients vital signs might be predictive for next event prediction.[GAP]

our model TEEDAM is a deep learning model for EHR that can simultanously 
encodes all of events and values and optimizes for a point process objective function as well as a desired outcome. We have performed extensive experiments on existing event sequence data to show the effectivness of our improved transformer event encoder against the existing methods. In addition, we show that encoding states can reduce point process NLL in two EHRs dataset (p12,p19) and in some cases it can lead to better performance for next event prediction. Finally, we provide explanations regarding the interpretablity of out model thanks to the utilized attention mechanisms.

the contributions of this paper are:
\begin{itemize}
    \item cont1
    \item cont2
\end{itemize}

explain structure of paper

@@

A temporal point process [13, 14] is a stochastic process that models a sequence of events in a continuous time domain. Consider an event sequence data  $\mathcal{D}=\{\mathcal{S}_{i}\}_{i=1}^N$ where $N$ is total number of samples and each sample $\mathcal{S}_{i}$ is represented as a sequence of events $\mathcal{S}_{i}=\{(t_i,e_i)\}_{j=1}^L$, where $L$ is the total number of occured events, $t_j$ is the event's timestamp, and $e_j \in \mathbb{R}^M $ is the binary represntation of event marks (multi-class or multi-label). Furthermore, the history of events at time $t$ is denoted as $\mathcal{H}_t=\{(t_j,e_j):t_j<t \}$.

The core idea of the point process framework is the definition of conditional intensity functions (CIFs) which is the probability of the occurence of an event of type $m$ in an infinitesimal time window $[t,t+dt)$:

\begin{equation} 
   \lambda_m^{*}(t)=\lim \frac{P(  \text{event of type m in } [t,t+\Delta t) |   \mathcal{H}_t  )}{\Delta t} 
\end{equation}

Here, $*$ denotes conditioning on the history of events. Multivariate Hawkes process is the traditional approach to characterize CIFs by assumming a fixed form of intensity to account for the additive influence of an history event:

@@

 where $\mu \ge 0 $, aka base intensity, is an exogenous component that is independent of history, while $\phi(t)>0$, excitation function, is an endogenous component depending on the history that shows the mutual influences. The excitation function can ge characterized using exponentials, linear combination of $M$ basis functions \cite*{xuLearningGrangerCausality2016}.

Encoder-decoder architectures have proven to be effective in many applications. The main idea of a neural temporal point process (NTPP) is to first encode the history of events until $t_j$ using a neural network architecture $h_j=Enc(\mathcal{H}_{j+1};\theta)$. Then it tries to estimate CIFs using a different decoder architectures $\lambda^*_m(t)=Dec(h_j;\phi)$ for $ t \in (t_j,t_{j+1}]$.

Initial works has used recurrent encoders such as RNN, GRU or LSTM [refs] in which the hidden state gets updated after arrival of new event as $h_{j+1}=Update(h_j,(t_j,e_j))$. The main advantage is that they allow us to compute history embeddings in $O(L)$ time, however, they are prone to neglect long-term inter-event influences. On the other hand, set aggregation encoders directly encodes all past events into a history embedding. adopting attention mechanism is one way that can capture long-term influences and can be trained in parallel as well. For example, \cite*{zuoTransformerHawkesProcess2020a} proposed transformer hawkes process (THP) that adopts a transformer architecture for event encoding.

The learned event embeddings can be later used for estimating conditional intensity functions, comulative conditional intensity functions or event probability density functions.

@@

Based on conditional intensity function, it is straightforward to derive conditional probability density function $p^{*}_{m}(t)$ in $(t_j, t_{j+1}]$:

\begin{equation} 
    p_m^{*}(t)=\lambda_m^{*}(t) \exp \left[-\sum_{m=1}^{M} \int_{t_j}^{t_{j+1}}\lambda_m^{*}(t^{\prime})dt^{\prime}\right]
 \end{equation}

 The paramters of point process can be learnt by Maximum Likeihood Estimation (MLE) framework. However, more advanced methods such as adversarial learning, reinforcement learning have also been proposed.

In the multi-class setting, the log-likelihood (LL) of a point process for a single event sequnce $\mathcal{S}_{i}$ is defined as:

\begin{multline} \label{eq:CIF-mc}
    \log p_{mc}(\mathcal{S}_{i})  =  \sum_{j = 1}^{L}\sum_{m = 1}^{M} \mathbbm{1}(e_j=m)   \log p_m^{*}(t_j) \\   
    = \sum_{j = 1}^{L}\sum_{m = 1}^{M} \mathbbm{1}(e_j=m)   \log \lambda_m^{*}(t_j)\\
    -\sum_{m = 1}^{M} \left(   \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right)
\end{multline}

Here, $\mathbbm{1}$ is the indicator function. The log-likelihood can be understood using the following two facts. First, the quantity $\lambda_k^{*}(t)dt$ corresponds to the probability of observing an event of type $k$ in the infinitesimal interval $[t_j,t_j+dt]$ conditioned on the past events $\mathbb{H}_{t_j}$. Second, we can compute the probability of not observing any eevents of type $k$ in the rest of the interval $[0,T]$ as $\exp \left(-\int_{0}^{T}  \lambda_m^{*}(s) ds\right)$.

However, in many cases such as EHRs it is common to have co-occuring events. \cite*{enguehardNeuralTemporalPoint2020} proposed to use a binary cross entropy function:

@@

It should be noted that the real advantage of point process is modeling non-event likelihoods in the form of integrals. If we neglect the integrals, we would achieve the cross-entropy and binary cross entropy loss in the multi-class and multi-label settings respectively for the prediction of next mark given history of events.

Another approach is the marked case which assumes that marks and timestamps are conditionally independent given $\mathcal{H}_t$.

\begin{multline} \label{eq:CIF-marked}
    \log p_{marked}(\mathcal{S}_{i}) =    
    \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log p^{*}(e_j=m) \\
    +\sum_{j = 1}^{L}\lambda^{*}(t_j) - \int_{t_0}^{t_{L}}\lambda^{*}(t^{\prime})dt^{\prime}  
\end{multline}

@@

@@

We focus on Neural point process based on attention mechanism that can potentially addresses the problems of slow serial computing and loss of long-term information. In addition, attention weights bring interpretability and can show peer influences between events. Self-attentive Hawkes Process (SAHP) \cite*{zhangSelfAttentiveHawkesProcess2020} proposes a multihead attention network as the history encoder. In addition, they use a sofplus function that can capture both excitaion and inhibition effects. Similarly, Transformer Hawkes Process (THP) \cite*{zuoTransformerHawkesProcess2020a} adopts the transformer architecture \cite*{vaswaniAttentionAllYou2017} with time encodings for event embeddings. They have introduced softplus function that can only capture mutual excitations between events. In a interesting study, reseraches studied different combination of encoders (self attention and GRU) and decoders and simulated on various datasets for comparison \cite*{enguehardNeuralTemporalPoint2020}. They demonsterated that attention-based TPPs appear to transmit pertinent EHR information and perform favourably compared to existing models. One gap in the current literature is that they don't have any mean for encoding additional information that can be useful for characterization of CIFs.

Recurrent neural networks have been modified to consider irregulary sampled time series. For example, GRU-D \cite*{cheRecurrentNeuralNetworks2018} adapts GRU to consider missingness pattern in the form of feature masks and time intervals to achieve better prediction results. RETAIN \cite*{choiRETAINInterpretablePredictive2017} is based on a two-level neural attention model that is specifically designed for EHRs data. SeFT \cite*{hornSetFunctionsTime2020} is based on recent advancements in differentiable set function learning, extremely parallelizable with a beneficial memory footprint, thus scaling well to large datasets of long time series and online monitoring scenarios. Their use of aggregation function is similar to Transformers that computes the embeddings of set elements independently, leading to lower runtime and memory complexity of O(n). Although, these models are nearly end-to-end that eliminates the need for imputation pipeline, it is still unclear how much they are affected by the missigness pattern in EHRs. In addition, they do not explicitly model the missingness pattern in the data.

@@

The proposed model, TEEDAM, consists of two modules for encoding a dataset with both event sequence data and irregularly sampled time series. The schematic of the model is depicted in \ref*{fig1}.

 The key advantage of our proposed model is to combine a transformer-based event encoder (TEE) with a deep attention module (DAM) that can handle an irregularly sampled time series for a any downstream prediciton task. The data is represented as $\mathcal{D}=\{(\mathcal{S}_i,\mathcal{U}_i)\}_{i=1}^{N}$, where $\mathcal{S}_i$ and $\mathcal{U}_i$ are event sequence data and irregular time series for the $i$-th sample.

\subsection{Event Encoder}
We use a transformer event encoder similar to THP \cite*{zuoTransformerHawkesProcess2020a} with minor modifications for event embedding.

In the first step, we embed all event marks $E_{emb}=E \times W_{emb}$ where $E_i \in \mathbb{R}^{L \times M}$ is the binary encoding matrix of all event marks (multi-label or multi-class), and  $ W_{emb} \in  \mathbb{R}^{M \times d_{emb} }  $ is the trainable embedding matrix. In addition, we encode vector of timestamps $t=[t_1,t_2,...,t_L] $ to $Z=[z(t_1),z(t_2),...,z(t_L)]\in \mathbb{R}^{L \times d_{time}}$ using following transformation formula:

@@

        \end{cases} 
\end{equation} 
Here, $\mathcal{T}$ is representing maximum time scale. This is very similar to positional encodings in transformers where the index is substitued by the time stamp. In contrary to THP and original positional encoding that considers $d_{emb}=d_{time}$ and adds up time encoding to the event embedding, we propose to concatenate this two vectors:

@@

Finally, we use the standard transformer encoder to encode embedded events $X_{emb}$ tothe encoded matrix  $ H=(h_1, ..., h_j, ..., h_L) $. It is important to use approporiate mask matrix to prevent information leakage from future to the past. In this case, $h_j$ containes the all available information until occurence of $j$-th event.

@@

We use the a deep attention module \cite*{hornSetFunctionsTime2020} for encoding all additional information including irregularly sampled time series. Each observation of a sample can be represented by $u_k=(z(t_k), v_k, m_k)$ where $z(t_k)$ is the same time encoding similar to equation \ref{eq:time-encoding}.% we define attention $a(\mathcal{U}_k,s_k )$

We define $\mathcal{U}_p$ to be the set of the first $p$ observations.
The goal is to calculate the attention weight $a(\mathcal{U}_p,u_k ), k \leq p$ that is the relevance of $k$-th observation $s_k$ to the first $p$ observed values $\mathcal{U}_p$.
This is achieved by computing an embedding of the set elements using a smaller set functions $f^{\prime}$, and projecting the concatenation of the set representation and the individual set element into d-dimensional space:

\begin{equation}\label{eq:fprime}
    y^{\prime}_{p}=f^{\prime}(\mathcal{U}_p) = g^{\prime} \left(  \frac{1}{|p|} \sum_{u_k \in \mathcal{U}_p}  h^{\prime}(u_k;\theta^{\prime}) ;\rho^{\prime}\right) 
\end{equation}

Here, we compute the mean of first $p$ observations after passing first $p$ observations through a multilayer perceptron (MLP) nueral network ($h^{\prime}(u_k;\theta^{\prime})$). Finally, a second transformation using $g^{\prime}$ is performed to obtain embeddings $f^{\prime}(\mathcal{U}_p) \in \mathbb{R}^{d_{g^{\prime}}}$.

@@

\begin{equation}
    a(\mathcal{U}_p,u_p)=  softmax(\frac{K_p.w^q}{\sqrt{d}  })
\end{equation}

@@

\subsection{All formulas}

@@

We perform various experiments to show the effectiveness of each component in out model.
\subsection*{Datasets}

>>>>

% https://github.com/James-Yu/LaTeX-Workshop/wiki/FAQ#how-to-pass--shell-escape-to-latexmk
\usepackage{svg}

@@

    Irregular sampling of time series in electronic health records (EHRs) presents a challenge for model development. In addition, the pattern of missingness for certain clinical variables is not random, but rather determined by clinicians' decisions and the patient's state. 
Point process is a mathematical framework for handling event sequence data that is consistent with irregular sampling. 
We propose \emph{NPP4EHR}, a deep neural network that can learn patient representations from irregularly sampled time series using the neural point process framework. We conducted various experiments to demonstrate the effectiveness of event and state encoding for characterizing conditional intensity functions, as well as for downstream prediction tasks, using two healthcare databases. 
Our proposed transformer event encoder outperforms state-of-the-art models in commonly used event sequence databases, as evidenced by its superior performance in negative log-likelihood and next-event prediction. Furthermore, we demonstrated the usefulness of state encoding for improved conditional intensity function characterization, as well as event encoding for better representation learning in downstream tasks, using two real-world healthcare datasets.
We propose that our model could be leveraged for better representation learning in EHRs.

\end{abstract}

@@

[ML in healthcare] Machine learning has the potential to revolutionize healthcare by leveraging vast amounts of data available in electronic health records (EHRs) to develop more accurate clinical decision support systems. EHRs store patient health information, such as medical history, medications, lab results, and diagnostic images, which can be used as input for machine learning algorithms to identify patterns and associations that could inform more precise diagnoses, better treatment plans, and earlier interventions. Clinical decision support systems that use machine learning can provide real-time, evidence-based recommendations to healthcare providers, reducing errors and improving patient outcomes.

[irregular sampling challenge] One of the data challenges for machine learning (ML) when using electronic health records (EHRs) is irregular sampling. EHR data is often collected at different times and frequencies, depending on a patient's healthcare needs and visit schedules, which can result in uneven and irregularly sampled time series data. Asynchronous and incomplete observation of certain clinical variables can also be regarded as missingness in the data.

@@

[sources of missingness] The sources of missing data in EHRs must be carefully understood. For instance, lab measurements are usually ordered as part of a diagnostic work-up, and the presence or absence of a data point conveys information about the patient's state \cite*{ghassemiReviewChallengesOpportunities2020}.

[imputation methods] Imputation techniques are widely used in electronic health records (EHRs) to handle missing data. Imputation refers to the process of filling in the missing data with plausible values based on the available information. There are several imputation techniques available, each with its strengths and weaknesses. For example, mean imputation replaces missing values with the mean of the available values in the same column, while regression imputation uses a regression model to estimate the missing values based on the relationship between the variables. Other popular imputation methods include multiple imputations, hot deck imputation, and k-nearest neighbors imputation. The choice of imputation technique depends on the type and amount of missing data, as well as the goals of the analysis. Regardless of the technique used, it is important to carefully consider the impact of imputed values on the analysis and to report the imputation method used in the results.

There exist several ML frameworks that are inherently compatible with the missing or irregularly-sampled data such as Gaussian process and recurrent neural networks.
Deep learning models that can handle irregularly sampled time series are crucial in many real-world applications, including stock market predictions, speech recognition, and medical diagnosis. These models must be able to process data that is not evenly spaced, which is common in time-sensitive applications. One popular deep learning architecture that can handle irregularly sampled time series is the Recurrent Neural Network (RNN). RNNs are well-suited for this task because they can process sequential data, taking into account not only the current input but also the previous inputs. 
Another architecture that can handle irregularly sampled time series is the Convolutional Neural Network (CNN), which can extract features from the data and then pass the processed data to an RNN for further analysis. Additionally, Attention Mechanisms can be integrated into RNNs and CNNs to help the model focus on important features in the data. These deep learning models offer a powerful toolset for handling irregularly sampled time series and providing useful insights into this type of data.
However, it is important to note that these models do not explicitly account for the fact that the absence of some data points can convey unique information, and it is important to carefully consider the sources of missingness in the data.

% Deep learning models that are compatible with irregularly sampled time series are important in many real-world applications, such as stock market predictions, speech recognition, and medical diagnosis. These models need to be able to handle data that is not evenly spaced, as is often the case in time-sensitive applications. One popular deep learning architecture that can handle irregularly sampled time series is the Recurrent Neural Network (RNN). RNNs are well-suited for this task because they can process sequential data, taking into account not only the current input but also the previous inputs. Another architecture that can handle irregularly sampled time series is the Convolutional Neural Network (CNN), which can be used to extract features from the data and then pass the processed data to an RNN for further analysis. Additionally, Attention Mechanisms can be integrated into RNNs and CNNs to help the model focus on important features in the data. Overall, these deep learning models offer a powerful toolset for handling irregularly sampled time series and provide useful insights into this type of data. However, these models do not explicitly take into account the fact wether the absence of some datapoints could convey some unique information.

Point process is the mathematical framework for describing the distribution of events in time or space. These events can be occurrences of earthquakes, nerve impulses, customer purchases, or anything else that can be counted or measured.
In healthcare, lab measurements can be regarded as sequence of events that are ordered by clinicians. At the core of point process is the defintion of conditional intensity functions and the corresponding log-likelihood which simultanously models the occurence of events using the history of past events.

Point process is a mathematical framework for describing the distribution of events in time or space. These events can be occurrences of earthquakes, nerve impulses, customer purchases, or anything else that can be counted or measured. In healthcare, lab measurements can be regarded as a sequence of events that are ordered by clinicians.
At the core of the point process is the definition of conditional intensity functions and the corresponding log-likelihood, which simultaneously models the occurrence of events using the history of past events. The conditional intensity function represents the instantaneous rate of event occurrence, given the history of past events. This function can be used to predict the probability of future events, which is particularly useful in healthcare applications such as predicting disease onset or detecting adverse events.

% More recently, neural Point Processes (NPP) have been developed for better characteriztion of CIFs by leveraing the power of deep neural networks. These models can be used for tasks such as predicting future events, estimating the rate of event occurrence, or identifying correlations between events. They offer a flexible and powerful way to analyze point process data, as they can handle complex dependencies between events and incorporate prior knowledge about the process. Traditioaly, a point process encodes timestamp and type of events, however, it is common in healtcare to have additional sources of information that could be useful for CIF characterization. For example the absolute value of patients vital signs might be predictive for next event prediction.

More recently, Neural Point Processes (NPP) have been developed to better characterize Conditional Intensity Functions (CIFs) by leveraging the power of deep neural networks. These models can be used for tasks such as predicting future events, estimating the rate of event occurrence, or identifying correlations between events. They offer a flexible and powerful way to analyze point process data, as they can handle complex dependencies between events and incorporate prior knowledge about the process.
Traditionally, a point process encodes the timestamp and type of events. However, in healthcare, additional sources of information can be available that could be useful for CIF characterization. For example, the absolute value of a patient's vital signs might be predictive for next-event prediction. NPP models can incorporate such information by using event and state encoders that can learn to extract relevant features from the input data.

Our proposed model, NPP4EHR, is a deep learning model for electronic health records (EHRs) that is based on point process theory. It consists of two modules: a transformer event encoder, which is designed to handle the missingness patterns of certain laboratory values, and a deep attention module for encoding all other available data. Both modules can be jointly learned by using appropriate loss functions for conditional intensity function (CIF) characterization and any downstream task.

We have conducted extensive experiments on existing event sequence data to demonstrate the effectiveness of our improved transformer event encoder in comparison to existing methods. Furthermore, we have shown that encoding states can reduce the negative log-likelihood (NLL) in two EHR datasets and, in some cases, lead to better performance for next event prediction. Overall, our proposed model offers a powerful tool for handling the irregularly sampled time series data present in EHRs and has the potential to improve representation learning in this domain.
The contributions of this paper are:
\begin{itemize}
    \item An improved transformer event encoder that can effectively handle missingness patterns in certain laboratory values and has been validated in baseline datasets.
    \item A deep attention module for encoding all available patient data, which has been shown to improve CIF characterization for laboratory values.
    \item Leveraging transformer event encoder in downstream tasks for better representation learning of patient's EHR
\end{itemize}

@@

A temporal point process is a stochastic process that models a sequence of events in a continuous time domain. Consider event sequence data  $\mathcal{D}=\{\mathcal{S}_{i}\}_{i=1}^N$ where $N$ is the total number of samples and each sample $\mathcal{S}_{i}$ is represented as a sequence of events $\mathcal{S}_{i}=\{(t_i,e_i)\}_{j=1}^L$, where $L$ is the total number of occurred events, $t_j$ is the event's timestamp, and $e_j \in \mathbb{R}^M $ is the binary representation of event marks (multi-class or multi-label). Furthermore, the history of events at time $t$ is denoted as $\mathcal{H}_t=\{(t_j,e_j):t_j<t \}$.

The core idea of the point process framework is the definition of conditional intensity functions (CIFs) which is the probability of the occurrence of an event of type $m$ in an infinitesimal time window $[t,t+dt)$:

\begin{equation} \label{eq:lambda_def}
   \lambda_m^{*}(t)=\lim \frac{P(  \text{event of type m in } [t,t+\Delta t) |   \mathcal{H}_t  )}{\Delta t} 
\end{equation}

Here, $*$ denotes conditioning on the history of events$\mathcal{H}_t$. The Multivariate Hawkes process is the traditional approach to characterize CIFs by assuming a fixed form of intensity to account for the additive influence of an event in the past:

@@

 where $\mu \ge 0 $, aka base intensity, is an exogenous component that is independent of history, while $\phi(t)>0$, excitation function, is an endogenous component depending on the history that shows the mutual influences. The excitation function can be characterized using exponentials, or linear combination of $M$ basis functions \cite*{xuLearningGrangerCausality2016}.

@@

 Based on the conditional intensity function \ref*{eq:lambda_def}, it is straightforward to derive conditional probability density function $p^{*}_{m}(t)$ in the interval $(t_j, t_{j+1}]$:

 \begin{equation} 
     p_m^{*}(t)=\lambda_m^{*}(t) \exp \left[-\sum_{m=1}^{M} \int_{t_j}^{t_{j+1}}\lambda_m^{*}(s)ds\right]
  \end{equation}

  The parameters of the point process can be learned by Maximum Likelihood Estimation (MLE) framework. However, more advanced methods such as adversarial learning, and reinforcement learning have also been proposed.

 In the multi-class setting, the log-likelihood (LL) of a point process for a single event sequence $\mathcal{S}_{i}$ is defined as:

 \begin{multline} \label{eq:CIF-mc}
     \log p_{mc}(\mathcal{S}_{i})  =  \sum_{j = 1}^{L}\sum_{m = 1}^{M} \mathbbm{1}(e_j=m)   \log p_m^{*}(t_j) \\   
     = \sum_{j = 1}^{L}\sum_{m = 1}^{M} \mathbbm{1}(e_j=m)   \log \lambda_m^{*}(t_j)\\
     -\sum_{m = 1}^{M} \left(   \int_{t_1}^{t_L}  \lambda_m^{*}(s) \,ds  \right)
 \end{multline}

 Here, $\mathbbm{1}$ is the indicator function. The log-likelihood can be understood using the following two facts. First, the quantity $\lambda_k^{*}(t)dt$ corresponds to the probability of observing an event of type $k$ in the infinitesimal interval $[t_j,t_j+dt]$ conditioned on the past events $\mathbb{H}_{t_j}$. Second, we can compute the probability of not observing any events of type $k$ in the rest of the interval $[t_1,t_L]$ as $\exp \left(-\int_{t_1}^{t_L}  \lambda_m^{*}(s) ds\right)$.

 However, in many cases, such as in EHRs, it is common to have co-occurring events. To handle this issue, \cite*{enguehardNeuralTemporalPoint2020} proposed using a binary cross-entropy function:

@@

%  It should be noted that the real advantage of point process is modeling non-event likelihoods in the form of integrals. If we neglect the integrals, we would achieve the cross-entropy and binary cross entropy loss in the multi-class and multi-label settings respectively for the prediction of next mark given history of events.

 It is important to note that the main advantage of point processes lies in their ability to model non-event likelihoods in the form of integrals. If we neglect the integrals, we would end up with the cross-entropy and binary cross-entropy loss in the multi-class and multi-label settings, respectively, for predicting the next mark given the history of events.

 Another approach is the marked case, which assumes that the marks and timestamps are conditionally independent given $\mathcal{H}_t$:

 \begin{multline} \label{eq:CIF-marked}
     \log p_{marked}(\mathcal{S}_{i}) =    
     \sum_{j = 1}^{L}\sum_{m = 1}^{M} \mathbbm{1}(e_j=m)   \log p^{*}(e_j=m) \\
     +\sum_{j = 1}^{L}\lambda^{*}(t_j) - \int_{t_1}^{t_{L}}\lambda^{*}(t^{\prime})dt^{\prime}  
 \end{multline}

@@

Encoder-decoder architectures have proven to be effective in many applications. The main idea of a neural temporal point process (NTPP) is to first encode the history of events until $t_j$ using a neural network architecture $h_j=Enc(\mathcal{H}_{j+1};\theta)$. Then it tries to estimate $m$ CIF using a different decoder architecture $\lambda^*_m(t)=Dec(h_j;\phi)$ for $ t \in (t_j,t_{j+1}]$.

% Initial works have used recurrent encoders such as RNN, GRU or LSTM [refs] in which the hidden state gets updated after the arrival of a new event as $h_{j+1}=Update(h_j,(t_j,e_j))$. The main advantage is that they allow us to compute history embeddings in $O(L)$ time, however, they are prone to neglect long-term inter-event influences. On the other hand, set aggregation encoders directly encode all past events into a history embedding. Adopting an attention mechanism is one such way that can capture long-term influences and can be trained in parallel as well which makes it more computationally efficient. For example, \cite*{zuoTransformerHawkesProcess2020a} proposed Transformer Hawkes Process (THP) that adopts a transformer architecture for event encoding.

Initial works have utilized recurrent encoders such as RNN, GRU, or LSTM, where the hidden state is updated after the arrival of a new event as $h_{j+1}=Update(h_j,(t_j,e_j))$. The main advantage of these models is their ability to compute history embeddings in $O(L)$ time. However, they are susceptible to ignoring long-term inter-event dependencies.
In contrast, set aggregation encoders encode all past events directly into a history embedding. One approach to capture long-term dependencies is to use an attention mechanism, which can be trained in parallel and is more computationally efficient. For instance,\cite{zuoTransformerHawkesProcess2020a} proposed Transformer Hawkes Process (THP), which adopts a transformer architecture for event encoding.

The learned event embeddings can be later used for estimating conditional intensity functions, cumulative conditional intensity functions or probability density functions.

@@

We focus on Neural point process models based on attention mechanisms that can potentially address the problems of slow serial computing and loss of long-term information. In addition, attention weights bring interpretability and can show peer influences between events. Self-attentive Hawkes Process (SAHP) \cite*{zhangSelfAttentiveHawkesProcess2020} proposes a multi-head attention network as the history encoder. In addition, they use a sofplus function that can capture both excitation and inhibition effects. Similarly, Transformer Hawkes Process (THP) \cite*{zuoTransformerHawkesProcess2020a} adopts the transformer architecture \cite*{vaswaniAttentionAllYou2017} with time encodings for event embeddings. They have introduced softplus function that can only capture mutual excitations between events. In an interesting study \cite*{enguehardNeuralTemporalPoint2020}, researchers studied different combinations of encoders (self-attention and GRU) and decoders and simulated on various datasets for comparison. They demonstrated that attention-based TPPs appear to transmit pertinent EHR information and perform favorably compared to existing models. One gap in the current literature is that they don't have any means for encoding additional information that can be useful for the characterization of CIFs. For example, in EHRs there are many information in addition to clinical events that could be useful

Recurrent neural networks have been modified to consider irregularly sampled time series. For example, GRU-D \cite*{cheRecurrentNeuralNetworks2018} adapts GRU to consider missingness patterns in the form of feature masks and time intervals to achieve better prediction results. RETAIN \cite*{choiRETAINInterpretablePredictive2017} is based on a two-level neural attention model that is specifically designed for EHRs data. SeFT \cite*{hornSetFunctionsTime2020} is based on recent advancements in differentiable set function learning, extremely parallelizable with a beneficial memory footprint, thus scaling well to large datasets of long time series and online monitoring scenarios. Their use of aggregation function is similar to transformers that compute the embeddings of set elements independently, leading to lower runtime and memory complexity of $O(n)$. Although these models are nearly end-to-end that eliminate the need for an imputation pipeline, it is still unclear how much they are affected by the missingness pattern in EHRs. In addition, they do not explicitly model the missingness pattern in the data.

@@

The proposed model, NPP4EHR, consists of two modules for encoding a dataset with both event sequence data and irregularly sampled time series. The schematic of the model is depicted in \ref*{fig1}.

 The key advantage of our proposed model is to combine a transformer-based event encoder (TEE) with a deep attention module (DAM) that can handle an irregularly sampled time series for any downstream prediction task. The data is represented as $\mathcal{D}=\{(\mathcal{S}_i,\mathcal{U}_i)\}_{i=1}^{N}$, where $\mathcal{S}_i$ and $\mathcal{U}_i$ are event sequence data and irregular time series for the $i$-th sample as described in previous section.

\subsection{Event Encoder}
We use a transformer event encoder similar to THP \cite*{zuoTransformerHawkesProcess2020a} with minor modifications for time encoding.
In the first step, we embed all event marks as $E_{emb}=E \times W_{emb}$, where $E_i \in \mathbb{R}^{L \times M}$ is the binary encoding matrix of all event marks (multi-label or multi-class), and  $ W_{emb} \in  \mathbb{R}^{M \times d_{emb} }  $ is the trainable embedding matrix. In addition, we encode vector of timestamps $t=[t_1,t_2,...,t_L] $ to $Z=[z(t_1),z(t_2),...,z(t_L)]\in \mathbb{R}^{L \times d_{time}}$ using following transformation formula:

@@

        \end{cases} 
\end{equation} 
Here, $\mathcal{T}$ is representing the maximum time scale. This transformation is very similar to positional encodings in transformers \cite*{vaswaniAttentionAllYou2017}, where the index is substituted by the timestamp. Contrary to THP and original positional encoding that considers $d_{emb}=d_{time}$ and adds up time encoding to the event embedding, we propose to concatenate these two vectors:

@@

Finally, we use the standard transformer architecture to encode embedded events $X_{ev}$ into the encoded matrix  $ H=(h_1, ..., h_j, ..., h_L) $. It is important to use an appropriate mask matrix to prevent information leakage from the future to the past. In this case, $h_j$ should contain all available information until the occurrence of $j$-th event.

@@

We use a deep attention module \cite*{hornSetFunctionsTime2020} for encoding all additional information including irregularly sampled time series. Each observation is represented by $u_p=(z(t_p), v_p, m_p)$ where $z(t_p)$ is the same transformation for time encoding in equation \ref{eq:time-encoding}.% we define attention $a(\mathcal{U}_k,s_k )$

We define $\mathcal{U}_p$ to be the set of the first $p$ observations.
The goal is to calculate the attention weight $a(\mathcal{U}_p,u_k ), k \leq p$ that is the relevance of $k$-th observation $u_k$ to the first $p$ observed values $\mathcal{U}_p$.
This is achieved by computing an embedding of the set elements using a smaller set functions $f^{\prime}$, and projecting the concatenation of the set representation and the individual set element into d-dimensional space:

\begin{equation}\label{eq:fprime}
   f^{\prime}(\mathcal{U}_p) = g^{\prime} \left(  \frac{1}{|p|} \sum_{u_k \in \mathcal{U}_p}  h^{\prime}(u_k;\theta^{\prime}) ;\rho^{\prime}\right) 
\end{equation}

Here, we compute the mean of the first $p$ observations after passing the first $p$ observations through a multilayer perceptron (MLP) neural network ($h^{\prime}(u_k;\theta^{\prime})$). Finally, a second transformation using $g^{\prime}$ is performed to obtain embeddings $f^{\prime}(\mathcal{U}_p) \in \mathbb{R}^{d_{g^{\prime}}}$.

@@

\begin{equation}
    a(\mathcal{U}_p,u_k)=  softmax(\frac{K_p.w^q}{\sqrt{d}  })
\end{equation}

@@

% \subsection{All formulas}

@@

We perform various experiments to show the effectiveness of each component in our model.
\subsection*{Datasets}
